---
title: "MA615 Final Project"
author: "Truc Minh Nguyen"
format: html
editor: visual
---


The Purpose of this report will focus on EDA and modelling a dataset
obtained regarding the Total capital stock for urban and rural areas in
the Turks and Caicos. Most of the Key Facts about TCI EDA is shown in the Shiny App and the Presentation for other Key Facts about TCI. 

```{r}
#| echo: false

library(tidyverse)
# install.packages("sf")
library(sf)
# install.packages("rnaturalearth")
# install.packages("rnaturalearthdata")
library(rnaturalearth)
library(rnaturalearthdata)
# install.packages("tmap", dependencies = TRUE)
library(tmap) # Optional for interactive maps
library(corrplot)
library(knitr)

```



# Let's retrieve spatial data for Turks and Caicos
```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")
turks_caicos <- world[world$name == "Turks and Caicos Is.", ]

# Let's plot a simple map of the Turks and Caicos Islands using ggplot

ggplot(data = turks_caicos) +
  geom_sf(fill = "lightblue", color = "black") +
  labs(
    title = "Map of Turks and Caicos Islands",
    caption = "Source: Natural Earth"
  ) +
  theme_minimal()


# Let's create and interactive map using tmap:

tmap_mode("view") # Set to interactive mode

tm_shape(turks_caicos) +
  tm_polygons(col = "lightblue", border.col = "black") +
  tm_layout(title = "Interactive Map of Turks and Caicos Islands")

```


# GDP Data of the Turks and Caicos
```{r}
GDP <- read.csv("Turks_GDP.csv", header = TRUE)
# Convert character column to Date format
GDP$Date <- as.Date(GDP$Date, format = "%m/%d/%Y")

# Extract only the Year
GDP$Year <- format(GDP$Date, "%Y")

GDP <- GDP %>%
  rename(
    GDP_billions_USD = GDP..Billions.of.US...,        
    Per_Capita = Per.Capita..US... 
  )
GDP$Per_Capita_Scaled <- GDP$Per_Capita / 10000  # Scale down for visualization


# Plot with adjusted y-axis limits
# Rescale Per Capita to match the GDP scale
GDP$Per_Capita_Scaled <- GDP$Per_Capita / 10000  # Adjust scaling

GDP_long <- GDP %>%
  pivot_longer(cols = c(GDP_billions_USD, Per_Capita_Scaled),
               names_to = "Metric", values_to = "Value")

# Updated Plot
GDP_plot <- ggplot(GDP_long, aes(x = Year, y = Value, group = Metric)) +
  # Bars for GDP
  geom_bar(data = subset(GDP_long, Metric == "GDP_billions_USD"),
           aes(fill = "GDP (Billions USD)"), 
           stat = "identity", alpha = 0.7) +
  # Line and points for Per Capita
  geom_line(data = subset(GDP_long, Metric == "Per_Capita_Scaled"),
            aes(color = "Per Capita (USD)"), size = 1) +
  geom_point(data = subset(GDP_long, Metric == "Per_Capita_Scaled"),
             aes(color = "Per Capita (USD)"), size = 2) +
  # Define the y-axis and secondary axis
  scale_y_continuous(
    name = "GDP (Billions USD)", 
    limits = c(0, 3.5),
    sec.axis = sec_axis(~ . * 10000, name = "Per Capita (USD)")
  ) +
  # Single legend with manual color and fill labels
  scale_fill_manual(name = "Metric", values = c("GDP (Billions USD)" = "skyblue")) +
  scale_color_manual(name = "Metric", values = c("Per Capita (USD)" = "red")) +
  
  # Labels and theme
  labs(
    title = "GDP (Billions USD) and Per Capita (USD) Over the Years",
    x = "Year",
    y = "GDP (Billions USD)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"  # Place legend at the top
  )

# Save the plot as a PNG image

# Save the plot with a white background
ggsave(
  filename = "GDP_plot_white_bg.png",
  plot = GDP_plot,
  width = 10, 
  height = 6,
  units = "in",
  dpi = 300,
  bg = "white"  # Explicitly set background to white
)

```


# Read in Capital Stock Data of TCI, create table of important variables and definitions
```{r}
TCI_stock <- read.csv("turks_model_df.csv", header = TRUE)

variable_table <- data.frame(
  "Important_Variables" = c("tot_cu","bed_prv_cu", "bed_pub_cu", "edu_prv_cu",
                            "edu_pub_cu", "emp_agr_cu", "emp_gov_cu", "emp_ind_cu",
                            "emp_ser_cu", "ic_high_cu", "ic_low_cu", "ic_mhg_cu",
                            "ic_mlw_cu", "tot_cr","bed_prv_cr", "bed_pub_cr",
                            "edu_prv_cr","edu_pub_cr", "emp_agr_cr",
                            "emp_gov_cr","emp_ind_cr","emp_ser_cr", "ic_high_cr", 
                            "ic_low_cr", "ic_mhg_cr","ic_mlw_cr"),
  "Definitions" = c(
    "Total capital stock urban (built environment) in million USD $",
    "Health-private sector-capital stock urban (built environment) in million USD $",
    "Health-public sector-capital stock urban (built environment) in million USD $",
    "Education-private sector-capital stock urban (built environment) in million USD $",
    "Education-public sector-capital stock urban (built environment) in million USD $",
    "Employment-agricol sector-capital stock urban (built environment) in million USD $",
    "Employment-government sector-capital stock urban (built environment) in million USD $",
    "Employment-industrial sector-capital stock urban (built environment) in million USD $",
    "Employment-service sector-capital stock urban (built environment) in million USD $",
    "Housing-high income group-capital stock urban (built environment) in million USD $",
    "Housing-low income group-capital stock urban (built environment) in million USD $",
    "Housing-upper middle income group-capital stock urban (built environment) in million USD $",
    "Housing-lower middle income group-capital stock urban (built environment) in million USD $",
    "Total capital stock rural (built environment) in million USD $",
    "Health-private sector-capital stock rural (built environment) in million USD $",
    "Health-public sector-capital stock rural (built environment) in million USD $",
    "Education-private sector-capital stock rural (built environment) in million USD $",
    "Education-public sector-capital stock rural (built environment) in million USD $",
    "Employment-agricol sector-capital stock rural (built environment) in million USD $",
    "Employment-government sector-capital stock rural (built environment) in million USD $",
    "Employment-industrial sector-capital stock rural (built environment) in million USD $",
    "Employment-service sector-capital stock rural (built environment) in million USD $",
    "Housing-high income group-capital stock rural (built environment) in million USD $",
    "Housing-low income group-capital stock rural (built environment) in million USD $",
    "Housing-upper middle income group-capital stock rural (built environment) in million USD $",
    "Housing-lower middle income group-capital stock rural (built environment) in million USD $"
    
  )
)

# Print the table
kable(variable_table, caption = "Important Variables and Definitions")

```



# Read in Capital Stock Data of TCI, look at correlation matrix :
We are interested in doing a linear regression to see what predictors can accurately
predict the Total Capital Stock for Urban and Rurul Locations in TCI. We will check 
assumptions to be able to fit a linear model. 
```{r}
urban_predictors <- TCI_stock[,c("tot_cu","bed_prv_cu", "bed_pub_cu", "edu_prv_cu",
                                 "edu_pub_cu", "emp_agr_cu", "emp_gov_cu", "emp_ind_cu",
                                 "emp_ser_cu", "ic_high_cu", "ic_low_cu", "ic_mhg_cu",
                                 "ic_mlw_cu")]
cor_matrix <- cor(urban_predictors, use = "pairwise.complete.obs")
urban_corrplot <- corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45)

#let's repeat this step for rural predictors
rural_predictors <- TCI_stock[,c("tot_cr","bed_prv_cr", "bed_pub_cr", "edu_prv_cr",
                                 "edu_pub_cr", "emp_agr_cr", "emp_gov_cr", "emp_ind_cr",
                                 "emp_ser_cr", "ic_high_cr", "ic_low_cr", "ic_mhg_cr",
                                 "ic_mlw_cr")]
cor_matrix2 <- cor(rural_predictors, use = "pairwise.complete.obs")
urban_corrplot <- corrplot(cor_matrix2, method = "color", type = "upper", tl.col = "black", tl.srt = 45)

```

From this step, it seems like everything is highly correlated with each other and 
there are some missing values. This is likely violating collinearity. But for the sake, 
let's entertain some models. 
```{r}
# Let's start with all the private sector predictors, with high income on the urban stock
urban_private_fit <- lm(tot_cu ~ bed_prv_cu + edu_prv_cu + emp_ind_cu + ic_high_cu, 
                        data = TCI_stock)
summary(urban_private_fit)

# Let's continue with all the private sector predictors, with low income on the urban stock
urban_public_fit <- lm(tot_cu ~ bed_pub_cu + edu_pub_cu + emp_ind_cu + ic_low_cu, 
                        data = TCI_stock)
summary(urban_public_fit)

# Let's predict urban stocks using rural predictors, let's keep bed_prv_cu because 
# significant predictor

urban_combine_fit2 <- lm(tot_cu ~ bed_prv_cu + bed_pub_cr + edu_pub_cr + emp_ind_cr 
                        + ic_high_cr + ic_low_cr,
                        data = TCI_stock)
summary(urban_combine_fit2)
```
It seems that the only significant predictors for the urban stock prediction model
with low p-value is the bed_prv_cu variable (Health-private sector-capital stock 
urban (built environment) in million USD $).Interesting to note that using rural 
predictors to estimate urban stock price results in non NA results. This is probably 
due to collinearity issue of using urban predictors to predict urban stocks.


Let's do the Rural Modeling using Urban Predictors: 
```{r}
rural_fit <- lm(tot_cr ~ bed_prv_cr + bed_pub_cu + edu_prv_cu + emp_ind_cu + 
                  ic_high_cu + ic_low_cu, 
                        data = TCI_stock)
summary(rural_fit)

```


When using urban predictors to predict rural stock, there are some NAs in model, so 
it seems that there is more collinearity issue. 

# Conclusion: 
Due to time constraints, EDA was not able to be performed thoroughly to choose 
the correct predictors and even the most appropriate model. Future works include 
spend more time on EDA, doing literature review. 

# References: 
This section includes all citations for sources used in the report, shiny app, and
presentation.

[1] Stanford University. (n.d.). [GAR15 Global Exposure Dataset for Turks and Caicos Islands]. 
Stanford Digital Repository. https://purl.stanford.edu/wb949hp4321
[2] Worldometer. (n.d.). Dominican Republic GDP. Worldometer. Retrieved June 17, 2024, from
https://www.worldometers.info/gdp/dominican-republic-gdp/
[3] Macrotrends. (n.d.). Cuba GDP per capita 1960-2024. Macrotrends. Retrieved June 17, 2024, from
https://www.macrotrends.net/global-metrics/countries/CUB/cuba/gdp-per-capita
[4] Worldometer. (n.d.). Bahamas GDP. Worldometer. Retrieved June 17, 2024, from
https://www.worldometers.info/gdp/bahamas-gdp/
[5] Government of Turks and Caicos Islands. (2023). National tourism development 
strategy and action plan 2023. Government of Turks and Caicos Islands. 
Retrieved June 17, 2024, from
https://gov.tc/tourismregulations/images/docs/TCI%20National%20Tourism%20Development%20Strategy%20and%20Action%20Plan%202023.pdf
[6] Government of Turks and Caicos Islands. (n.d.). Turks and Caicos Islands statistics department. Government of Turks and Caicos Islands. Retrieved June 17, 2024, from https://www.gov.tc/stats/
[7] Government of Turks and Caicos Islands. (n.d.). Cabinet. Government of Turks and Caicos Islands. Retrieved June 17, 2024, from https://www.gov.tc/premier/cabinet
[8] Government of Turks and Caicos Islands. (n.d.). Tourism statistics. 
Government of Turks and Caicos Islands. Retrieved June 17, 2024, from https://www.gov.tc/stats/statistics/economic/41-tourism
[9] Encyclopaedia Britannica. (n.d.). Caribbean Sea. Encyclopaedia Britannica. 
Retrieved June 17, 2024, from https://www.britannica.com/place/Caribbean-Sea
[10] Central Intelligence Agency. (n.d.). Turks and Caicos Islands: Flag. 
The World Factbook. Retrieved June 17, 2024, from
https://www.cia.gov/the-world-factbook/countries/turks-and-caicos-islands/flag/





